---
title: Economic Activity in Canada
description: A map of economic activity in Canada as represented by the location and number of bank branches
author: Bruce Meng
date: '2017-11-13'
slug: map-branch
categories: []
tags:
  - data
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(leaflet)
library(leaflet.extras)
library(tidyverse)

branch <- read_csv("../../data/branches/fib.data.big6.csv")
```

```{r data.process, include = F}
branch.clean <- select(branch, Bank.name, lon, lat) %>%
                        .[complete.cases(.),]


```

# Description

Despite the advance of mobile and digital banking, the physical branch is still a preferred place to do banking business for many consumers and service businesses. By looking at where and how many bank branches are, we can get a glimpse of where economic activity is in Canada.

This is a map of branch locations from the major Canadian banks.

Branch locations from:

* Bank of Montreal
* Scotiabank
* Canadian Imperial Bank of Commerce
* National Bank
* Royal Bank of Canada
* Toronto-Dominion Bank

Date of data:

> Sep. 2017

Source of data:

> [Payments Canada](https://www.payments.ca/our-directories/financial-institutions-branch-directory). Geocoded by [Photon](http://photon.komoot.de/).

# Map

```{r map, message = F, echo = F, warning = F}
bank.colr <- colorFactor(c("#0000FF", "#FF0000", "#8B6508", "#00CD00"), c("BMO", "CIBC", "TD", "RBC"))

leaflet(branch.clean) %>%
        addProviderTiles(provider = "CartoDB.DarkMatter") %>%
        # addCircleMarkers(lng = branch.clean$lon,
        #            lat = branch.clean$lat,
        #            stroke = F, 
        #            color = ~bank.colr(branch.clean$Bank),
        #            group = branch.clean$Bank,
        #            label = paste(branch.clean$Bank, branch.clean$Address, sep = ": ")
        #            ) %>%
        addHeatmap(layerId = "Heat On", group = "Heat On",
                   radius = 5) %>%
        # addLayersControl(overlayGroups = branch.clean$Bank,
        #                  baseGroups = c("Heat On", "Heat Off")) %>%
        addFullscreenControl()
        # addLegend(pal = bank.colr,
        #           values = ~branch.clean$Bank,
        #           position = "bottomright",
        #           title = "")

```

## Observations

Most of the economic activity is located near the border to the U.S. (not surprising). It did surprise me on how sparse the economic activity is from Northern Saskatchewan all the way to Northern Quebec. Conversely, I was also surprised that there is significant enough activity as far north as Cambridge Bay!

## Other

I developed something similar at work, but instead of viewing it from a lens of where economic activity is, I viewed it from a lens of where the competition is. I didn't reproduce individual bank branch locations here in hopes of keeping that slight edge in competitive intelligence ;).