---
title: Market Simulation Model
description: A model to simulate the behaviour of equity markets
author: Bruce Meng
date: '2017-10-22'
showdate: false
slug: market-simulation-model
categories: []
tags: ["modelling"]
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(forecast)
library(magrittr)
library(readxl)
library(readr)
library(ggplot2)
library(dplyr)
library(ggthemes)
library(plotly)
library(scales)
set.seed(1)
source("../../data/simul.model/Simul.model.v2.GLD.R")
```


## Description

This model will generate reasonable prices for the market and will use the latest real-time price as the starting point.

Date updated: `r Sys.time()`

## TSX

The Toronto Stock Exchange (TSX) historical and simulated values are modelled below:

```{r tsx, echo = F, fig.width = 7, fig.height = 5, message = F, warning = F}

# Length of historical TSX to plot
data.TSX.plot <- filter(data.TSX, Date > 2007)

# Simulation date
date.simul <- as.Date(mc.TSX$Date[1]) %>%
                as.numeric()

# Plot simulation
TSX.g <- ggplot(data = mc.TSX, aes(x = as.Date(Date)))

# Historical plot
TSX.g <- TSX.g +geom_line(data = data.TSX.plot, aes(x = as.Date(Date), y = GSPTSE.Close))

# Simulation plot
for (paths in 1:paths) {
    TSX.g <- TSX.g +geom_line(aes_string(y = names(mc.TSX)[paths]), alpha = 0.17, colour = "blue")
}

# Simulation Avg
TSX.g <- TSX.g +geom_line(aes(y = Avg), linetype = "dashed")

TSX.g <- TSX.g +geom_vline(xintercept = date.simul, linetype = "dotted")

TSX.g <- TSX.g +ggtitle("TSX historical + simulation", subtitle = paste("Expected path of the TSX, based on past behaviour and", paths, sep = " ", "runs")) +
                ylab("TSX Index") +
                xlab("Date") +
                scale_y_continuous(labels = comma)

                

TSX.g <- TSX.g +theme_minimal()
ggplotly(TSX.g)
```

## S&P 500

The Standard & Poor's 500 (S&P 500) index historical and simulated values are modelled below:

```{r sp500, echo = F, fig.width = 7, fig.height = 5, message = F, warning = F}
# Length of historical SP500 to plot
data.SP500.plot <- filter(data.SP500, Date > 2007)

# Simulation date
date.simul <- as.Date(mc.SP500$Date[1]) %>%
                as.numeric()

# Plot simulation
SP500.g <- ggplot(data = mc.SP500, aes(x = as.Date(Date)))

# Historical plot
SP500.g <- SP500.g +geom_line(data = data.SP500.plot, aes(x = as.Date(Date), y = GSPC.Close))

# Simulation plot
for (paths in 1:paths) {
    SP500.g <- SP500.g +geom_line(aes_string(y = names(mc.SP500)[paths]), alpha = 0.17, colour = "blue")
}

# Simulation Avg
SP500.g <- SP500.g +geom_line(aes(y = Avg), linetype = "dashed")

SP500.g <- SP500.g +geom_vline(xintercept = date.simul, linetype = "dotted")

SP500.g <- SP500.g +ggtitle("S&P500 historical + simulation", subtitle = paste("Expected path of the S&P500, based on past behaviour and", paths, sep = " ", "runs")) +
                        ylab("S&P500 Index") +
                        xlab("Date") +
                        scale_y_continuous(labels = comma)
        

SP500.g <- SP500.g +theme_minimal()
ggplotly(SP500.g)
```

## How it works

The model builds the simulation based on historical logarithmic returns of each market. It utilizes the generalized lambda probability density function as the underlying model to draw returns from. A more detailed explanation can be found in Posts.
