---
title: '"Drop the Mic" machine learning (part 2)'
author: Bruce Meng
date: '2018-01-14'
slug: drop-the-mic-machine-learning-part-2
categories: []
tags:
  - R
  - modelling
  - fun
---

I wrote about an 'algorithm' that I concocted in my head for determining "Drop the Mic" winners in an earlier post: ["Drop the MIC" winner algorithm!](https://www.brucemeng.ca/post/drop-the-mic/). In that post I found that my silly little algorithm has accurately predicted the winner 85% of the time. `r emo::ji("thumbsup")`

To rehash that algorithm I said:

1. If one contestant is female and the other is male, the female wins!
2. If both are male, or both are female, or if they are a group, the one standing to the left of the screen at the end wins!

That was fun, but it wasn't very data science-y. So... let's apply a more proper machine learning method to determine the winner algorithm.

We start again with scraping the data from IMDb:

```{r data}
library(rvest)
library(tidyverse)
library(knitr)
library(kableExtra)

# Scrape data from IMDb
data.dropmic <- read_html("http://www.imdb.com/title/tt6396094/episodes?season=1&ref_=tt_eps_sn_1") %>%
        html_nodes("strong a") %>%
        html_text() %>%
        as.data.frame() %>%
        .[grepl("vs", .[,1]),] %>%
        data.frame(Episode = .)

data.clean <- str_split(data.dropmic$Episode, "&| and ") %>%
        unlist() %>%
        data.frame(Battles = .)

# Display results
kable(data.clean, "html") %>%
        kable_styling(bootstrap_options = c("striped", "hover"))
```

The Battles were properly scraped in the above table. Let's grab the following features from it:

1. Who are the contestants?
2. What is the gender of contestant 1 (we shall designate the first name in the battle as contestant 1)?
3. Where is contestant 1 standing (left or right on the screen)?

```{r features.1}
# One data cleaning step needs to be done... make all "vs. " to "vs "
data.clean$Battles <- gsub("vs. ", "vs ", data.clean$Battles)

## 1. Who are the contestants?
# Separate contestants
features <- str_split(data.clean$Battles, "vs ")

# Use purrr::map to add Contestants to data.clean
data.model <- cbind(data.clean, 
                    Contestant.1 = purrr::map(features, 1) %>% unlist(),
                    Contestant.2 = purrr::map(features, 2) %>% unlist())

# Display results
kable(data.model, "html") %>%
        kable_styling(bootstrap_options = c("striped", "hover"))
```

We have our contestants! Let's now add the genders of both contestants (I'm doing this manually... training a convolutional neural network to recognize females is a bit out of scope for this post `r emo::ji("stuck_out_tongue")`) as well as the positions and winners:

```{r features.2, echo = F, message = F}
data.model <- read_csv("../../data/dropmic/data.model.csv")
data.model$Con1.gender <- factor(data.model$Con1.gender)
data.model$Con2.gender <- factor(data.model$Con2.gender)
data.model$Con1.position <- factor(data.model$Con1.position)
data.model$winner <- factor(data.model$winner)

data.model.comp <- data.model[complete.cases(data.model),]

# Display results
kable(data.model, "html") %>%
        kable_styling(bootstrap_options = c("striped", "hover"))

```

Nice! We are now ready to train a machine learning model to predict the winner:

```{r caret, message = F}
library(caret)

# Create training dataset
train.index <- createDataPartition(data.model.comp$winner, 
                                   p = 0.8,
                                   list = F)

data.train <- data.model.comp[train.index,]
data.test <- data.model.comp[-train.index,]

## Establish caret parameters
# Training controls
caret.train.control <- trainControl(
        method = "repeatedcv",
        number = 10,
        repeats = 10
        )
# Training machine learning model
doParallel::registerDoParallel(cores = 6)

machine.model <- train(winner ~ Con1.gender + Con2.gender + Con1.position, 
                       data = data.train, 
                       method = "rf",
                       trControl = caret.train.control)

# Model details
machine.model
```

The model, based on training data, is reporting a 71% accuracy. But the real test is on the test dataset, `data.test`:

```{r caret.test}
# Predicting winners on test dataset
mm.predict <- predict(machine.model, newdata = data.test)

results <- cbind(data.test, mm.predict)

# Display results
kable(results, "html") %>%
        kable_styling(bootstrap_options = c("striped", "hover"))

```

The machine learning algorithm predicted that contestant 1 would win all 3 battles. It was right 2/3 times.

And now finally... the main battle!

When I made part 1, the results of the last two battles were not released yet. Machine vs. Human... Joshua, give me a beat!

Human prediction (based on algo developed in part 1):

1. Danielle Fishel will beat Jonathan Lipnicki (female)
2. Shania will beat Meghan

```{r battle}
data.battle <- data.model[(nrow(data.model) - 1): nrow(data.model),]

# Machine prediction
mm.predict <- predict(machine.model, data.battle)

results <- cbind(data.battle, mm.predict)

# Display results
kable(results, "html") %>%
        kable_styling(bootstrap_options = c("striped", "hover"))
```

Machine prediction:

1. Danielle Fishel will beat Jonathan Lipnicki
2. Meghan will beat Shania

(I was a bit worried that the machine learning algo would make the same prediction as me... that would've been quite... anti-climactic)

## Battle 1: Danielle Fishel vs. Jonathan Lipnicki

<center><iframe width="560" height="315" src="https://www.youtube.com/embed/XxTjaF02u7E" frameborder="0" allow="autoplay; encrypted-media" allowfullscreen></iframe></center><br>

And the winner is Danielle! Both human and machine score a point!

## Battle 2: Tie-breaker: Shania Twain vs. Meghan Trainor

<center><iframe width="560" height="315" src="https://www.youtube.com/embed/_ckccU8ojRg" frameborder="0" allow="autoplay; encrypted-media" allowfullscreen></iframe></center><br>

